---
title: "WHO MAKES 100K+ IN TORONTO:EVIDENCE FROM SEPTEMBER 2023 LABOUR FORCE SURVEY"
author: "Gaurja Newatia"
date: "`r Sys.Date()`"
output: pdf_document
warnings: False
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(broom)
library(caret)
library(pROC)
library(stargazer)
library(fastDummies)
library(tidyr)
library(knitr)
library(kableExtra)
library(naniar)
library(corrplot)
library(vcd)
library(forcats)
library(DataExplorer)
set.seed(123)
```

### Loading the Data

```{r include=FALSE}
setwd("C:/Users/Gaurja Newatia/Documents/GitHub/Toronto-Wages-Modelling")
data <- read_excel("LFS.xlsx",sheet = "data")
new_names<-tolower(names(data))
names(data)<-new_names
```

### 1. Introduction

The world is currently grappling with a global recession that has left
no country untouched,including Canada. Despite nominal wage increases,
Canada faces the harsh reality that these wage gains have not kept pace
with the rising rate of inflation.Moreover,with millions of people
immigrating to Canada every year, the labour supply has drastically
increased, which has profound implications on wage growth. In response
to these challenges, this paper aims to uncover the key factors
influencing wages in Canada and seeks to provide policymakers with
valuable insights that can inform the development of policies promoting
wage equity.

For the purpose of our research, we restrict our analysis to Toronto,
one of Canada’s largest Census Metropolitan Area (CMA), with population
of 6.6 million residents. The problem of interest is to classify the
surveyed population as:

-   High Earners (Earning an Annual Gross Salary of \$100,000 or More)
-   Non-High Earners (Earning an Annual Gross Salary of less than
    \$100,000)

### 2. Summary of Previous Work

A lot of research has been conducted in the past and continues to be
ongoing, to enable policymakers formulate policies that govern wage
determination and empower them address wage differentials.

Mueller, R.E. (2022) analyzed the merged monthly Labour Force Survey
data to investigate the gender paygap across four distinct public and
private sector definitions, as well as within each of these sectors.
They argue that when compared across gender, females tend to exhibit
higher wage premiums in the public sector compared to their male
counterparts. Moreover, the gender paygap is consistently in favour of
males within all sectors, with the disparity slighlty larger in the
private sector.

Using 42 years of Current Population Survey Data (1977-2018), McConnell,
et al (2023) present a novel finding: marriage has a positive causal
effect on the wages of both men and women. They argue that there is a
significant shift for married women, as the once-existent wage penalty
associated with marriage has transformed into a wage premium. They
attribute the premium for men can be predominantly attributed to
household specialization, akin to Becker's theory (1985, 1993), where
married men tend to focus on market work, resulting in increased market
hours and higher earnings compared to their single counterparts.
Historically, the wage penalty for women also stemmed from household
specialization, with married women working considerably fewer market
hours than singles. However, over time, decreased specialization between
spouses has enabled married women to nearly match the work hours of
singles, eliminating the penalty.

The Quality of Employment Report of Statistics Canada (2021) gives some
insights on the state of hourly wages in Canada. It argues that even
though average wages in Canada is gradually rising since 1988, certain
sections of employees continued to have lower hourly wages than others
in 2021. It emphasizes that being a younger employee (aged 15 to 24),
having a high school diploma or less, and working part-time were all
factors associated with lower hourly wages.

### 3. Description of the Data

The Labour Force Survey (LFS) is a monthly cross-sectional survey
conducted by Statistics Canada which are used to produce the
unemployment rate as well as other standard labour market indicators
such as the employment rate and the participation rate. The monthly
dataset used in this study pertains to September 2023 with a 90%
response rate.

By applying a filter to the Labour Force Survey (LFS) specific to the
Toronto Census Metropolitan Area (CMA), we've narrowed down our sample
to encompass 5,645 individuals. This dataset comprises information on 60
variables, collectively addressing a wide range of subjects, including:

-   Employment and Unemployment

-   Industries

-   Labour

-   Occupations

-   Unionization and industrial relations

-   Wages, salaries and other earnings

The 'Earnings' dummy variable, which is our response variables allows us
to categorize individuals based on their income level.We classify
individuals as 'High Earners' if their hourly wage is greater than or
equal to 48.08 [^1], and as 'Non High Earners' if their hourly wage is
less than 48.08 CAD. This classification is based on the assumption that
someone working 40 hours a week at a wage of 48.08 CAD per hour or
higher would earn an annual gross salary of \$100,000 or more.

[^1]: The data is taken from <https://ca.talent.com/convert>

### 4. Description of Data Issues

[**4.1 Missing Values**]{.underline}

```{r Dropping Variables, include=FALSE}
drop_cols<-c("everwork" ,"ftptlast","prevten","durunemp" ,"flowunem", "unemftpt","whylefto", "whyleftn" ,"durjless", "availabl" ,"lkpubag","lkemploy", "lkrels", "lkatads","lkansads", "lkothern" ,"prioract" ,"ynolook","tlolook","survyear" ,"survmnth","rec_num", "lfsstat" ,"prov", "cma" )

data_new<-drop_columns(data = data,drop_cols)



```

```{r include=FALSE}
numerical<-c("wksaway","uhrsmain","ahrsmain","utothrs","atothrs","hrsaway","paidot","unpaidot","xtrahrs","tenure","hrlyearn","finalwt")


numeric_data<-data_new[numerical]
categorical<-data_new[setdiff(names(data_new), numerical)]

categorical_variables<-read_excel("categorical_variables.xlsx",sheet="EDA Sheet")

numeric_data$atothrs<-numeric_data$atothrs/10
numeric_data$utothrs<-numeric_data$utothrs/10
numeric_data$ahrsmain<-numeric_data$ahrsmain/10
numeric_data$uhrsmain<-numeric_data$uhrsmain/10
numeric_data$hrsaway<-numeric_data$hrsaway/10
numeric_data$paidot<-numeric_data$paidot/10
numeric_data$unpaidot<-numeric_data$unpaidot/10
numeric_data$xtrahrs<-numeric_data$xtrahrs/10
numeric_data$hrlyearn<-numeric_data$hrlyearn/100

combined_data<-cbind(categorical_variables,numeric_data)
combined_data["earnings_dummy"]<-data_new$`earnings dummy`


```

To facilitate our analysis, we organized the dataset into numerical and
categorical variables. Upon reviewing the dataset, we noticed the
presence of missing values within several numerical variables, even
though these missing values hold specific interpretations. For example,
in the "paidot" variable, which represents the number of paid overtime
hours, blank values signify that an individual does not receive paid
overtime. To address this issue and ensure the inclusion of these
missing values in our dataset, we made the decision to impute them with
the value "999999." This imputation process was carried out using
Microsoft Excel.

Similarly for categorical variables, we opted to impute missing values
with "Not Applicable." This choice was made to enable the incorporation
of these variables in our analysis, even when specific categorical
information was absent.

Table 1 below describes the missing values after imputation

```{r Missing Values, echo=FALSE}

missing_summary <- miss_var_summary(data)

filtered<-missing_summary[missing_summary$pct_miss>0,]
table2<-kable(filtered,col.names = c("Variable name","Number of Missing Values","Missing Values(%)"),align = "c",caption = "Percentage of Missing Values")
table1 <- kable_styling(table2, full_width = F) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1:3, border_left = TRUE, border_right = TRUE)
table1
```

Table 1 containes categorical variables and those with 100% missing
values. Consequently, we made the decision to remove variables that
exhibited 100% missing values. Additionally, we opted to exclude
"rec_num," "lfsstat," "prov," and "cma" from our dataset. These
variables primarily serve as indicators in our filtered dataset and do
not contribute meaningful information to our analysis.

```{r include=FALSE}
drop_cols<-c("everwork" ,"ftptlast","prevten","durunemp" ,"flowunem", "unemftpt","whylefto", "whyleftn" ,"durjless", "availabl" ,"lkpubag","lkemploy", "lkrels", "lkatads","lkansads", "lkothern" ,"prioract" ,"ynolook","tlolook","survyear" ,"survmnth","rec_num", "lfsstat" ,"prov", "cma" )

data_new<-drop_columns(data = data,drop_cols)


data_new <- data_new %>%
  mutate_all(~ifelse(is.na(.),"Not Applicable",.))

data_new_temp<-data_new[data_new$ftptmain!=2,]
```

### 5. Exploratory Data Analysis

[**5.1 Distribution of High and Non-High Earners**]{.underline}

The distribution of high earners and non-high earners is indicated in
Figure:1

```{r include=FALSE}
percentage_data <- combined_data %>%
  group_by(earnings_dummy) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)

```

```{r Distribution of High and Non-High Earners, echo=FALSE, message=FALSE, warning=FALSE}


ggplot(data = percentage_data, aes(x=earnings_dummy, y=Percentage)) +
  labs(x = "Earnings", y = "Percentage") +
  geom_text(aes(label = paste0(round(Percentage, 2), "%")), vjust = -0.5) +
  geom_bar(stat="identity", fill="lightblue") +
  scale_x_continuous(breaks = c(0, 1)) + # Setting the breaks for the x-axis
  theme(plot.title = element_text(size = 12, hjust = 0.5))





```

[**5.2 Differences in Earning Group**]{.underline}

While the average hourly wage rates in the Canadian labor market have
experienced a gradual increase, it's important to note that certain
groups of employees still face challenges, resulting in lower hourly
wage rates. This can be observed by the relatively lower proportion of
individuals categorized as "High Earners" in the subsequent sections.

[5.2.1 Based on Demographic Characteristics]{.underline}

Figure 2 reveals that women clearly have lower proportion of High
Earners than men in Toronto. Furthermore, being a younger employee (aged
15-19) or having less than a bachelor's degree significantly impacts the
likelihood of falling into the "High Earner" category. These findings
are concisely summarized in Figure 3 and Figure 4, respectively.

Figure 5 shows interesting results: Married individuals exhibit a
considerably higher representation among high earners compared to other
marital categories. Additionally, Figure 6 reveals that immigrants
constitute a smaller proportion of high earners when compared to
non-immigrants. This phenomenon may be attributed to a substantial
portion of immigrants in our dataset who engage in part-time employment
and are international students, which leads to reduced hourly earnings.

```{r Differences in Earnings, echo=FALSE, message=FALSE, warning=FALSE}
percentage_sex <- combined_data %>%
  group_by(earnings_dummy,sex) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=sex))+
  facet_wrap(~sex)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 0.9), size = 3) +  
  labs(title = "Figure:2 Percentage of High and Non-High Earners by Gender", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 8,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_sex

percentage_age <- combined_data %>%
  group_by(earnings_dummy,age_12) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=age_12))+
  facet_wrap(~age_12)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 0.9), size = 3) +  
  labs(title = "Figure:3 Percentage of High and Non-High Earners by Age Group", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 8,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_age


percentage_marstat <- combined_data %>%
  group_by(earnings_dummy,marstat) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=marstat))+
  facet_wrap(~marstat)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 0.9), size = 3) +  
  labs(title = "Figure:4 Percentage of High and Non-High Earners by Marital Status", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 8,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_marstat


percentage_educ <- combined_data %>%
  group_by(earnings_dummy,educ) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=educ))+
  facet_wrap(~educ)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 0.9), size = 3) +  
  labs(title = "Figure:5 Percentage of High and Non-High Earners by Education", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 5,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_educ

percentage_immig <- combined_data %>%
  group_by(earnings_dummy,immig) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=immig))+
  facet_wrap(~immig)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 0.9), size = 3) +  
  labs(title = "Figure:6 Percentage of High and Non-High Earners by Immigration", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 5,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_immig


  






```

[5.2.2 Based on Sector]{.underline}

Figure 7 reveals that while private sector has a significantly higher
proportion of high earners than the public sector, it also has a higher
proportion of "Non-High Earners" perhaps due to higher number of
employment opportunities offered by this sector at lower wages.

```{r echo=FALSE, message=FALSE, warning=FALSE}

percentage_sector <- combined_data %>%
  group_by(earnings_dummy,cowmain) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=cowmain))+
  facet_wrap(~cowmain)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 1), size = 3) +  
  labs(title = "Figure 7: Percentage of High and Non-High Earners by Sector", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 5,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_sector

```

Figure 8 indicates that an establishment with huge number of employees
pays its employees well as compared to smaller establishments. Though, a
huge proportion of non-high earners are in establishments with 20-99
employees, this could be due to the fact that smaller companies often
hire part time workers, particularly students which reduced their wage
costs.

```{r echo=FALSE, message=FALSE, warning=TRUE}
percentage_est <- combined_data %>%
  group_by(earnings_dummy,estsize) %>%
  summarise(Percentage = n() / nrow(combined_data) * 100)%>%
  ggplot()+
  geom_col(aes(x=earnings_dummy,y=Percentage,fill=estsize))+
  facet_wrap(~estsize)+
  
  geom_text(aes(x = earnings_dummy, y = Percentage, label = paste0(round(Percentage,2), "%"), vjust = 1.5), size = 3) +  
  labs(title = "Figure 8: Percentage of High and Non-High Earners by Establishment Size", x = "Earnings", y = "Percentage")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(size = 5,face = "bold"))+
  theme(legend.position = "none")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

percentage_est
```

[**5.3 Distribution of Hourly Earnings**]{.underline}

The gross hourly earnings (before taxes and other deductions) has a
right skewed distribution, i.e. a higher proportion of individuals are
in the lower tail, as indicated by Figure:9

```{r Hourly Earnings Distribution, echo=FALSE}

uhrlyearn<-ggplot(data = combined_data,aes(x=hrlyearn))+
geom_density(color="lightpink")+
labs(title = "Figure 9:Density of Hourly Earnings", x = "Hourly Earnings")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))

uhrlyearn



```

[**5.3 Distribution of Total Hours Worked**]{.underline}

Figure 10 indicates that most people in Toronto work on an average 40
hour per week, which reveals a good work life balance.

```{r echo=FALSE}
utothrs_worked<-ggplot(data = numeric_data,aes(x=utothrs))+
geom_density(fill="lightyellow")+
labs(title = "Figure 10: Density of Usual Total Hours Worked", x = "Usual Total Hours Worked")+
  theme(plot.title = element_text(size = 12,hjust = 0.5))
utothrs_worked
```

[**5.4 Exploring thought provoking hypothesis**]{.underline}

In this section, we seek to validate intriguing findings presented in
existing literature through straightforward data analysis and
exploration.

[5.4.1 Does marriage lead to wage premium?]{.underline}

Inspired by the insights of McConnell, et al (2023), we aim to address
this question for Toronto CMA.

Figure 11 and Figure 12 summarize the hourly earnings of females and
males respectively based on their marital status. It is interesting to
note that married females and males have higher median earnings as
compared to their single counterparts, which could possibly be explained
by the notion that married people can wait longer for better paying
jobs, since they have their partner's income to support themselves
during the job search period.

```{r echo=FALSE}
females<-subset(combined_data,sex=="Female")

males<-subset(combined_data,sex=="Male")
combined_sex<-rbind(females,males)

ggplot(data = females,aes(x=marstat,y=hrlyearn,fill=marstat))+
geom_boxplot()+
labs(title = "Figure 11: Box Plot showing wage differentials for Females Based on Marital Status", x = "Marital Status")+
theme(plot.title = element_text(size = 10,hjust = 0.5))+
theme(legend.position = "none")+
theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(data = males,aes(x=marstat,y=hrlyearn,fill=marstat))+
geom_boxplot()+
labs(title = "Figure 12: Box Plot showing wage differentials for Males Based on Marital Status", x = "Marital Status")+
theme(plot.title = element_text(size = 10,hjust = 0.5))+
theme(legend.position = "none")+
theme(axis.text.x = element_text(angle = 45, hjust = 1))




```

[5.4.1 Do females earn better in the public sector?]{.underline}

Inspired by Mueller, R.E. (2022), we aim to address this question for
Toronto CMA.

Figure 13 and Figure 14 show a higher proportion of females in the left
tail of the distribution in both the private and the public sector.
However, in the public sector, the proportion of females earning low
wages than males is slightly lower than in the case of private sector.

```{r echo=FALSE}
public<-subset(combined_data,cowmain=="Public sector employees")

ggplot(data = public,aes(x=hrlyearn,color=sex))+
geom_density()+
labs(title = "Figure 13: Density Plot of Earnings of Males and Females in the Public Sector", x = "Hourly Earnings")+
theme(plot.title = element_text(size = 10,hjust = 0.5))
```

```{r echo=FALSE}
private<-subset(combined_data,cowmain=="Private sector employees")
ggplot(data = private,aes(x=hrlyearn,color=sex))+
geom_density()+
labs(title = "Figure 14: Density Plot of Earnings of Males and Females in the Private Sector", x = "Hourly Earnings")+
theme(plot.title = element_text(size = 10,hjust = 0.5))
```

As observed in Figure 15, the proportion of females in the left tail of
the distribution is higher in the private sector than in the public
sector. The same is true for males, as indicated in Figure 16. This
gives us some evidence of lower wage disparity in the public sector.

```{r echo=FALSE}
public_female<-subset(public,sex=="Female")
private_female<-subset(private,sex=="Female")
combined_female<-rbind(public_female,private_female)
ggplot(data =combined_female,aes(x=hrlyearn,color=cowmain))+
geom_density()+
labs(title = "Figure 15: Density Plot of Earnings of Females in the Private and Public Sector", x = "Hourly Earnings",fill="Sector")+
theme(plot.title = element_text(size = 10,hjust = 0.5))


```

```{r echo=FALSE}
public_male<-subset(public,sex=="Male")
private_male<-subset(private,sex=="Male")
combined_male<-rbind(public_male,private_male)
ggplot(data =combined_male,aes(x=hrlyearn,color=cowmain))+
geom_density()+
labs(title = "Figure 16: Density Plot of Earnings of Males in the Private and Public Sector", x = "Hourly Earnings")+
theme(plot.title = element_text(size = 10,hjust = 0.5))
```

### 6. Variable Selection

The analysis accounts for several individual and job characteristics.
Age of the individual is characterized using indicators for five-year
age groups. Education is characterized to show the impact of college
education on earning capabilities i.e. we divide the individuals as
”College Education” and ”No-College Education” (set equal to ”College”
if the person has ”Above bachelor’s degree”,”Postsecondary certificate
or diploma” or a ”Bachelor’s degree and ”No-College” otherwise ). An
indicator variable for marital status is set equal to one if the
individual reports being legally married or living in a commonlaw union
and zero otherwise. For union status, we create an indicator for whether
a person is covered by a collective agreement (set equal to one if the
individual is a member of a union or covered by the collective
agreement, zero otherwise). Furthermore, to study the influence of
having children, we use the ”ageyonk” variable and divide the
individuals as ”No-Children” if variable ageyonk is equal to
Not-Applicable and ”Has-Children” otherwise. we also study the impact of
overtime eligibility on earnings capability by distinguishing the
individuals as overtime-eligible if they get paidot and not-eligible
otherwise. A variable for tenure represents the number of months a
person has held his or her current job and the variable utothrs
indicates the usual total hours that a person works at their main job.
Finally, we separate the individuals based on the industry they are
working in i.e. we call them working in a ”Top-5” industry if they work
in ”Real estate and rental and leasing”,”Finance and insurance”,
”Manufacturing - non-durable goods”,”Manufacturing -durable goods” and
”Health care and social assistance” and ”Not Top-5” otherwise.

```{r include=FALSE}
num_data_final<-drop_columns(data = numeric_data,c("hrlyearn","finalwt","uhrsmain","ahrsmain","atothrs","hrsaway","wksaway","unpaidot","xtrahrs"))

```

```{r include=FALSE}
categorical_variables["Children"]<-ifelse(categorical_variables$agyownk=="Not applicable","No Children","Has Children")
categorical_variables["Overtime Eligible"]<-ifelse(num_data_final$paidot==99999.9,"Not Eligible","Eligible")
top5_industries<-c("Real estate and rental and leasing","Finance and insurance",
"Manufacturing - non-durable goods","Manufacturing -durable goods","Health care and social assistance")

categorical_variables["Industry"]<-ifelse(categorical_variables$naics_21 %in% top5_industries, "Performers", "Non-Performers")

marriage<-c("Married", "Living in common-law")
categorical_variables["Marital_Status"]<-ifelse(categorical_variables$marstat %in% marriage, "Married/Common-Law", "Single")

College<-c("Above bachelor's degree","Postsecondary certificate or diploma","Bachelor's degree" )

categorical_variables["College"]<-ifelse(categorical_variables$educ %in% College, "College Education","No College Education")

Non_Union<-c("Non-unionized")
categorical_variables["Union_Status"]<-ifelse(categorical_variables$union %in% Non_Union,"Non Unionized Job","Unionized Job")


categorical_variables_final<-drop_columns(categorical_variables,c("age_6","noc_10"  , "noc_43", "yabsent", "payaway" , "yaway", "whypt","schooln",  "efamtype" ,"agyownk","firmsize","naics_21","marstat","educ","union"))
num_data_final<-drop_columns(num_data_final,"paidot")

combined_wd<-cbind(num_data_final,categorical_variables_final)

combined_wd<-combined_wd[combined_wd$ftptmain!="Part-time",]
combined_wd["Earnings_Dummy"]<-data_new_temp$`earnings dummy`
combined_wd<-combined_wd[combined_wd$age_12!="15 to 19 years",]


dummy_colnames<-c("age_12", "sex","College"  , "mjh" , "cowmain" ,  "immig" ,     "Union_Status", "permtemp","estsize", "Children",  "Overtime Eligible", "Industry","Marital_Status")

numeric_colnames<-c("utothrs","tenure","Earnings_Dummy")


```



```{r include=FALSE}
## Creating Dummy Variables
dummy_cat<-dummy_cols(combined_wd[dummy_colnames],remove_selected_columns=T,remove_first_dummy = T)
dummy_cat<-lapply(dummy_cat, as.factor)
#cols_drop<-c("age_12_25 to 29 years","sex_Male","marstat_Married","educ_High school graduate","cowmain_Public sector employees" ,"immig_Non-immigrant","union_Union member","permtemp_Permanent","estsize_More than 500 employees" , "Industry_Non-Performers" ,"Children_No Children", "Overtime Eligible_Eligible","Marital_Status")
#dummy_cat<-drop_columns(dummy_cat,cols_drop)

combined_final<-cbind(combined_wd[numeric_colnames],dummy_cat)

combined_final$tenure<-scale(combined_final$tenure)
combined_final$utothrs<-scale(combined_final$utothrs)
```

```{r include=FALSE}
### Dividing the data into train(80%) and test (20%)
index <- createDataPartition(combined_final$Earnings_Dummy, p = 0.8, 
                             list = FALSE)
train_data <- combined_final[index, ]
test_data <- combined_final[-index, ]
```

```{r include=FALSE}
custom_labels <- c("Usual hours","Tenure","Earnings_Dummy", "25 to 29 years",                             
  "30 to 34 years" ,                            
 "35 to 39 years"  ,                           
  "40 to 44 years" ,                            
 "45 to 49 years" ,                            
 "50 to 54 years" ,                            
"55 to 59 years"  ,                           
 "60 to 64 years" ,                            
 "65 to 69 years" ,                            
"70 and over", "sex_Male","No College","Single jobholder, including job changers","Public sector employees"   ,"Self-employed incorporated, ","Immigrant, landed more than 10 years earlier"
, "Non-immigrant" ,"Unionized Job","Temporary, casual or other temporary jobs"
, "Temporary, seasonal job"                  
, "Temporary, term or contract job"  ,"estsize_100 to 500 employees"                      
, "estsize_Less than 20 employees"                    
,"estsize_More than 500 employees" ,"No Children" ,"Overtime Not Eligible", "Industry_Performers" ,                              
 "Marital_Status_Single"   )

colnames(combined_final)<-custom_labels

```

### 7. Fitting the Logistic Regression Model

Since the outcome of interest is a binary variable, a binary logistic
regression model with a logit link was fit. Variable Selection was done
as indicated. Next, we address the issue with having varying ranges
amongst our numerical variables by standardizing them so that they have
a mean of 0 and standard deviation of 1.This is done to ensure that the
scales of the variables does not impact the contribution of the variable
to the outcome of interest. we then proceed to build three models:

-   Model 1 includes all the covariates
-   Model 2 includes all variables from Model 1 in addition to
    interaction effects between marital status and sex as well as sector
    where an individual is working and sex. These interaction effects
    are based on review of literature
-   In Model 3, we drop the insignificant variables obtained in Model 2
    to create Model

For the three explanatory models, we use the entire dataset, since
reducing the sample induces bias but for the predictive models, we
divide the dataset into train (80%) and test (20%) as the goal of
predictive modelling is to sacrifice bias in order to reduce sampling
variance, that is key to improving generalization (model performance on
test dataset). The predictive models are furthermore generated via 10
fold repeated cross validation of train data to ensure the robustness
and reliability of the model’s performance estimates. By repeatedly
dividing the data into 10 different subsets and using each subset in
turn for validation while training on the remaining data, we were able
to mitigate the risk of overfitting and gain a more accurate
understanding of how the model is likely to perform on unseen data.
Finally,to generate predictions, we use the threshold of 0.50 to get
binary outcomes.

Fitting Explanatory Models

Model-1

```{r message=FALSE, warning=FALSE}
fit_ogl<-glm(Earnings_Dummy~., data = combined_final, family = binomial(link = "logit"))

```

Model-2

```{r}
fit<-glm(Earnings_Dummy ~ `Usual hours` + Tenure + `25 to 29 years` + 
    `30 to 34 years` + `35 to 39 years` + `40 to 44 years` + 
    `45 to 49 years` + `50 to 54 years` + `55 to 59 years` + 
    `60 to 64 years` + `65 to 69 years` + `70 and over` + 
    sex_Male + `No College` + `Single jobholder, including job changers` + 
    `Public sector employees` + `Self-employed incorporated, ` + 
    `Immigrant, landed more than 10 years earlier` + `Non-immigrant` + 
    `Unionized Job` + `Temporary, casual or other temporary jobs` + 
    `Temporary, seasonal job` + `Temporary, term or contract job` + 
    `estsize_100 to 500 employees` + `estsize_Less than 20 employees` + 
    `estsize_More than 500 employees` + `No Children` + 
    `Overtime Not Eligible` + Industry_Performers + Marital_Status_Single+
    Marital_Status_Single*sex_Male+sex_Male*`Public sector employees`
    ,data = combined_final,family =binomial(link = "logit"))
```

```{r}
fit_reduced<-glm(Earnings_Dummy~`Usual hours` + Tenure + `25 to 29 years` + 
 `30 to 34 years` + `35 to 39 years` + `40 to 44 years` + `45 to 49 years` + 
   `50 to 54 years` + `55 to 59 years` + `60 to 64 years` + `65 to 69 years` + 
   `70 and over` + sex_Male + `No College` + `Public sector employees` +  
   `Non-immigrant` + `Unionized Job` + `estsize_100 to 500 employees` 
 + `estsize_Less than 20 employees` +  `estsize_More than 500 employees`
 + `No Children` + Marital_Status_Single+
   sex_Male*`Public sector employees` 
 ,data =  combined_final,family =binomial(link = "logit"))

```

Fitting Predictive Models

Model-1

```{r message=FALSE, warning=FALSE}
cv <- trainControl(method = "repeatedcv", number = 10,repeats = 10)
model_cv_ogl<-train(as.factor(Earnings_Dummy)~., data=train_data, method = "glm",
family="binomial",trControl = cv)
```

Model-2

```{r message=FALSE, warning=FALSE}
cv <- trainControl(method = "repeatedcv", number = 10,repeats = 10)
model_cv_1 <- train( as.factor(Earnings_Dummy)~utothrs + tenure + 
      `age_12_25 to 29 years` + `age_12_30 to 34 years` + `age_12_35 to 39 years` + 
    `age_12_40 to 44 years` + `age_12_45 to 49 years` + 
    `age_12_50 to 54 years` + `age_12_55 to 59 years` + 
    `age_12_60 to 64 years` + `age_12_65 to 69 years` + 
    `age_12_70 and over` + sex_Male + `College_No College Education` + 
    `mjh_Single jobholder, including job changers` + `cowmain_Public sector employees` + 
    `cowmain_Self-employed incorporated, with paid help` + 
    `immig_Immigrant, landed more than 10 years earlier` + 
    `immig_Non-immigrant` + `Union_Status_Unionized Job` + 
    `permtemp_Temporary, casual or other temporary jobs` + 
    `permtemp_Temporary, seasonal job` + `permtemp_Temporary, term or contract job` + 
    `estsize_100 to 500 employees` + `estsize_Less than 20 employees` + 
    `estsize_More than 500 employees` + `Children_No Children` + 
    `Overtime Eligible_Not Eligible` + Industry_Performers + 
    Marital_Status_Single+ Marital_Status_Single*sex_Male+
      sex_Male*`cowmain_Public sector employees`, 
    data = train_data, method = "glm", family="binomial",trControl = cv)
```

Model-3

```{r message=FALSE, warning=FALSE}
cv <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
model_cv_2 <- train(as.factor(Earnings_Dummy)~utothrs + tenure + 
    `age_12_25 to 29 years` + 
    `age_12_30 to 34 years` + `age_12_35 to 39 years` + 
    `age_12_40 to 44 years` + `age_12_45 to 49 years` + 
    `age_12_50 to 54 years` + `age_12_55 to 59 years` + 
    `age_12_60 to 64 years` + `age_12_65 to 69 years` + 
    `age_12_70 and over` + sex_Male + `College_No College Education` + 
     `cowmain_Public sector employees` + 
    `immig_Non-immigrant` + `Union_Status_Unionized Job` + 
    `estsize_100 to 500 employees` + `estsize_Less than 20 employees` + 
    `estsize_More than 500 employees` + `Children_No Children` + 
    Marital_Status_Single+sex_Male*`cowmain_Public sector employees`, 
    data = train_data, method = "glm", family="binomial",trControl = cv)
```

### 8. Model Performance

Model Validation in explanatory modelling involves goodness of fit-tests
while in predictive modelling, the focus is on comparing a model’s
generalization abilities. We therefore use the Likelihood-Ratio Test
(LRT) to evaluate model fit and the model test accuracy and the Expected
Prediction Error (EPE) to compare models’ predictive abilities.

The LRT method, in simple terms compares the log likelihood ratio
statistic from the two nested models to a Chi-squared distribution. This
determines whether it is beneficial to add more parameters or if the
simpler model is preferred.

For predictive power, using 10 fold repeated cross validation, we train
Model 1, Model 2 and Model 3 on the train data and evaluate their
predictive accuracy on the test data using EPE. The EPE is the average
of the difference between predicted values and actual values.

Model Validation for Explanatory Models

```{r echo=FALSE}
LM1<-lmtest::lrtest(fit_ogl,fit_reduced,fit)
LM1
```

Model with interaction terms and with interaction terms+ insignificant
variables shows a better fit.

Model Validation for Predictive Models

Results for 10 fold repeated cross validation

```{undefined echo=FALSE}
results <- resamples(list(Model_ogl=model_cv_ogl,Model_1 = model_cv_1,
                          Model_2 = model_cv_2))
s_results<-summary(results)
s_results
```

```{r}
predictions_model_cv_ogl<- predict(model_cv_ogl, newdata = test_data, 
                                   type = "prob")

predictions_model_cv_1<- predict(model_cv_1, newdata = test_data,
                                 type = "prob")

predictions_model_cv_2<- predict(model_cv_2, newdata = test_data,
                                 type = "prob")
```

```{r}
test_probs_cv_ogl <- predictions_model_cv_ogl[, "1"]
test_probs_cv_1 <- predictions_model_cv_1[, "1"]
test_probs_cv_2 <- predictions_model_cv_2[, "1"]
```

Calculation of Brier Score

```{r}
mean((test_probs_cv_ogl-test_data$Earnings_Dummy)^2)
mean((test_probs_cv_1-test_data$Earnings_Dummy)^2)
mean((test_probs_cv_2-test_data$Earnings_Dummy)^2)
```

When evaluating the models via LR test, we find Model 2 and Model 3 have
significant Chisq values at 1% significance level, that indicates Model
2 and Model 3 are a better fit when compared to Model 1 and Model 3
performs even better than Model 2 in terms of model fit and in addition,
it contains all the significant covariates.

On the contrary, Model 1 has the lowest Brier Score among all the three
models. Hence, it would be reasonable to argue that while focusing on
explanatory power, since we need to maximize model fit,we should go with
Model 3 but when the target is to maximise predictive accuracy, Model 1
would be an appropriate choice.

### 9. AUC/ROC Curves

```{r}
roc_object<-roc(test_data$Earnings_Dummy, test_probs_cv_ogl,percent=T)
plot(roc_object,legacy.axes=T,percent=T,xlab="False Positive Percentage",
     ylab="True Positive Percentage",lwd=4,col="blue")
text(0.7, 0.2, paste("AUC =", round(auc(roc_object), 2)), col = "blue")

```

```{r}
roc_object<-roc(test_data$Earnings_Dummy, test_probs_cv_1,percent=T)
plot(roc_object,legacy.axes=T,percent=T,xlab="False Positive Percentage",
     ylab="True Positive Percentage",lwd=4,col="blue")
text(0.8, 0.2, paste("AUC =", round(auc(roc_object), 2)), col = "blue")

```

```{r}
roc_object<-roc(test_data$Earnings_Dummy, test_probs_cv_2,percent=T)
plot(roc_object,legacy.axes=T,percent=T,xlab="False Positive Percentage",
     ylab="True Positive Percentage",lwd=4,col="blue")
text(0.8, 0.2, paste("AUC =", round(auc(roc_object), 2)), col = "blue")
```

For the test data, Area under the curve is the highest for Model-1 when
compared to Model-2 and Model-3.

### 10. Confusion Matrix

```{r}
test_probs_cv_ogl_class<-ifelse(test_probs_cv_ogl>0.50,1,0)
test_probs_cv_1_class<-ifelse(test_probs_cv_1>0.50,1,0)
test_probs_cv_2_class<-ifelse(test_probs_cv_2>0.50,1,0)
```

```{r}
CM1<-confusionMatrix(as.factor(test_data$Earnings_Dummy), 
                     as.factor(test_probs_cv_ogl_class))

CM1
```

```{r}
CM2<-confusionMatrix(as.factor(test_data$Earnings_Dummy), 
                     as.factor(test_probs_cv_1_class))
CM2
```

```{r}
CM3<-confusionMatrix(as.factor(test_data$Earnings_Dummy),
                     as.factor(test_probs_cv_2_class))

CM3
```

Based on the confusion matrices for the three models-

-Sensitivity Model-2\>Model-3\>Model-1 -Specificity Model-2\> Model-3\>
Model-1

Clearly, Model-2 Outperforms the other two models when it comes to
sensitivity (True Positive Rate) and specificity(True Negative Rates).

Thus, while Model-1 gives us highly accurate predictions (lowest Brier
Score), it however fails to capture True Positive and True Negatives in
the model. This could be due the fact that our dataset is severely
imbalanced and has a high proportion of low earners as compared to high
earners.
